---
title: "Compare Groups + Vizzz"
output:
  github_document:
    toc: true
author: prise6, `r format(Sys.Date(), "%d/%m/%Y")`
always_allow_html: yes
editor_options: 
  chunk_output_type: console
---


### Motivations

I found a really great package recently called __[compareGroups](http://www.comparegroups.eu/)__ that build bivariate tables. I think it's a great work and i want to go further in the rendering.

According to my short researchs, this package have been existing since [2014](https://www.jstatsoft.org/article/view/v057i12), but is recently out on [github](https://github.com/isubirana/compareGroups) and the [CRAN](https://cran.r-project.org/web/packages/compareGroups/index.html).

V4 is currently only available on github:

```{r, eval=FALSE}
library(devtools)
devtools::install_github(repo = "isubirana/compareGroups")
```

_(this package has many dependences)_


This _tutorial_ is not a copy of the [wonderfull vignette](https://cran.r-project.org/web/packages/compareGroups/vignettes/compareGroups_vignette.html) of the package because it's really well explained. Again, great job, really useful !


### Goal

Let's say you have a clustering to perform, and you'd like to explain your clusters with figures and graphs. This will be a dump example with the ... `iris` dataset. This post does not focus on clustering, it's only a pretext to have this final rendering:

```{r, echo=FALSE}
readRDS(file.path(getwd(), "Datas", "01-comparegroups-plots.rds"))
```


Let's see the step we're going through

1. Packages
2. Data and global variables
3. Perform quick clustering with PCA and K-means
4. Compare clusters
5. __Compare clusters with graphs and figures__


### 1. Packages

We need to load and attach few packages.

* `compareGroups`: THE package we want to _hack_
* `xml2`: DOM manipulation
* `htmltools`: html tags manipulation
* `knitr`: dynamic report generation
* `ggplot2`: draw some graphs


```{r, message=FALSE}
library(compareGroups)
library(xml2)
library(htmltools)
library(knitr)
library(kableExtra)
library(ggplot2)
library(data.table)
```


### 2. Data and global variables

I set differents variables that we need along the script. Obvisously i set le `NB_CLUSTERS` to 3 because it's not the point of this post to find the best number of clusters.

```{r global_variables}
# global variables
VARIABLES             = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")
VARIABLES_AND_CLUSTER = c("cluster", VARIABLES)
NB_CLUSTERS           = 3
NSTART                = 100
SEED_KM               = 234
```

We fake that `iris` dataset doesn't contain any group and we want to cluster observations. We create `iris.x` datatable on which we'll perform the analysis.

```{r data_prep, collapse=TRUE}
# load dataset
data(iris)
# make it datatable
setDT(iris)
# create iris.x
iris.x = iris[, ..VARIABLES]
str(iris.x)
```


### 3. Perform quick clustering with PCA and K-means

Before computing K-means i'd like to project `iris.x` to one or two dimension space thanks to PCA. 

```{r pca}
# PCA scaled
PCA = prcomp(iris.x, scale. = T, center = T)
# add PCA components to iris.x datatable
iris.x = cbind(iris.x, as.data.table(PCA$x))
```

```{r, collapse=TRUE}
# see that the two first component explains majority of variability
cumsum(PCA$sdev)/sum(PCA$sdev)
```


We chose two dimensions to compute K-means with `r NB_CLUSTERS` distinct centers and test `r NSTART` intials configurations.

```{r kmeans}
# reproducibility
set.seed(SEED_KM)
# K-means
model.km = kmeans(iris.x[, list(PC1, PC2)], centers = NB_CLUSTERS, nstart = NSTART)
# add cluster columng to iris.x datatable
iris.x$cluster = model.km$cluster
# transform into factor with nice labels
iris.x[, cluster := factor(
  x      = cluster,
  levels = seq_len(NB_CLUSTERS),
  labels = paste("Cluster", seq_len(NB_CLUSTERS))
)]
```


### 4. Compare clusters

Let's use `compareGroups` to compute mean of the observations foreach variables and clusters.

```{r compare_group_1}
comparegroups.main = compareGroups(
  formula          = cluster ~ .,
  data             = iris.x[, ..VARIABLES_AND_CLUSTER]
)
```

```{r, collapse=TRUE}
comparegroups.main
```


Above, we print the summary of the future table we're creating. Notice that `p.value` are computed, if distribution is normal, anova is performed. Otherwise you'll find great explanations on [this vignette](https://cran.r-project.org/web/packages/compareGroups/vignettes/compareGroups_vignette.html).

Next, we create our table, with a _ALL_ column, meaning all our observations.

```{r compare_group_2}
comparegroups.main.table = createTable(
  x        = comparegroups.main,
  show.all = T
)
```

```{r, collapse=TRUE}
comparegroups.main.table
```


Then, we can render this table into markdown format thanks to `export2md()` function using knitr. We store the result in `comparegroups.html` variable to modify content later.

```{r compare_group_3}
comparegroups.html = suppressWarnings(
  export2md(
    x             = comparegroups.main.table,
    caption       = "",
    header.labels = c(
      "all"       = "All",
      "p.overall" = "p-value"
    )
  )
)
```

```{r, collapse=TRUE}
comparegroups.html
```


### 5. Compare clusters with graphs and figures

First, we need to make small plots. Plots are images that we need to encode into [base64 string](https://en.wikipedia.org/wiki/Base64). The idea is to store the image inside the HTML thanks to [data URI scheme](https://en.wikipedia.org/wiki/Data_URI_scheme)

For example, this HTML code is an `html <img>` tag sourcing (`src` attribut) a __base64 image__ thanks to __URI scheme__:

```{html, eval=FALSE}
<img src="data:image/png;base64,iVBORw0KGgoAAA
ANSUhEUgAAAAUAAAAFCAYAAACNbyblAAAAHElEQVQI12P4
//8/w38GIAXDIBKE0DHxgljNBAAO9TXL0Y4OHwAAAABJRU
5ErkJggg==" alt="Red dot" />
```

In your browser this HTML is a red dot:

<img src="data:image/png;base64,iVBORw0KGgoAAA
ANSUhEUgAAAAUAAAAFCAYAAACNbyblAAAAHElEQVQI12P4
//8/w38GIAXDIBKE0DHxgljNBAAO9TXL0Y4OHwAAAABJRU
5ErkJggg==" alt="Red dot" />


#### Function to generate plot

`generatePlot()` asks for the selected cluster and the whole dataset containing `PC1` and `PC2`. It returns a `ggplot2` object with two `geom_point` layer. In black, the non selcted clusters. In red, the selected cluster. `PC1` as x-axis and `PC2` as y-axis.

```{r generatePlot}
generatePlot = function(cluster_s, data) {
  general_graph_data = data[!cluster %in% cluster_s, list(PC1, PC2)]
  cluster_graph_data = data[cluster %in% cluster_s, list(PC1, PC2)]
  
  g = ggplot() +
    geom_point(data = general_graph_data, aes(x = PC1, y = PC2), alpha = .3) +
    geom_point(data = cluster_graph_data, aes(x = PC1, y = PC2), color = "red") +
    theme_void()
  
  return(g)
}
```

Example:

```{r}
(p = generatePlot("Cluster 1", iris.x))
```


#### Function to generate data URI scheme

We need to save our plot as an image, read it thanks to `knitr::image_uri()` function that create data URI scheme.

`plotToURI()` asks for the ggplot object to transform and additionnal arguments for `ggsave()` function.

```{r plotToURI}
plotToURI = function(plot, file = NULL, ...) {
  
  if(is.null(file)) file = paste(tempfile(), "png", sep = ".")
  
  ggsave(filename = file, plot = plot, ...)
  uri = knitr::image_uri(file)
  
  return(uri)
}
```

```{r}
pURI = plotToURI(p, width = 3, height = 3, units = "cm", dpi = 100)
```

See what is like.

```{r, collapse=TRUE}
str(pURI, nchar.max = 80)
```

If we use `img` tag with the `htmltools` package around that string we observe this image:

```{r}
tags$img(src = pURI)
```


#### HTML Tables

Our table is made up of HTML tags. Rows are coded with `<tr>` tag and we want to add a new row where i wrote the comment `<!-- plot row -->`


```{html}
<table>
<caption></caption>
 <thead>
  <tr>
   <th style="text-align:left;">   </th>
   <th style="text-align:center;"> All </th>
   <th style="text-align:center;"> Cluster 1 </th>
   <th style="text-align:center;"> Cluster 2 </th>
   <th style="text-align:center;"> Cluster 3 </th>
   <th style="text-align:center;"> p-value </th>
  </tr>
 </thead>
<tbody>
  <!-- plot row -->
  <tr>
   <td style="text-align:left;">  </td>
   <td style="text-align:center;"> N=150 </td>
   <td style="text-align:center;"> N=53 </td>
   <td style="text-align:center;"> N=47 </td>
   <td style="text-align:center;"> N=50 </td>
   <td style="text-align:center;">  </td>
  </tr>
  ...
</tbody>
</table>
```

So we need to add to the HTML table a block like :

```{html}
<tr>
  <td style="text-align:left;"></td>
  <td style="text-align:left;"></td>
  <td style="text-align:center;">
    <img src="data:image/png;base64,..."/>
  </td>
  <td style="text-align:center;">
    <img src="data:image/png;base64,..."/>
  </td>
  <td style="text-align:center;">
    <img src="data:image/png;base64,..."/>
  </td>
  <td style="text-align:left;"></td>
</tr>
```

Let's create this html block by looping on clusters.

```{r}
# create img tags and td tags around
html_block = lapply(paste("Cluster", 1:3), function(cl) {
  p = generatePlot(cl, iris.x)
  uri = plotToURI(p, width = 3, height = 3, units = "cm", dpi = 100)
  
  tags$td(style = "text-align:center;", tags$img(src = uri))
})

# fill the block with td tags in empty cells
html_block = tagList(
  tags$td(style = "text-align:left;"),
  tags$td(style = "text-align:left;"),
  html_block,
  tags$td(style = "text-align:left;")
)

# we wrap the block in a tr tag
html_block = tags$tr(html_block)
```


#### Parse HTML

Ok, we did the hardest part. Now, we want to combine the table and the html block. This is where we need `xml2` package to parse XML/HTML elements.

We have to create a XML document instance thanks to `read_html()`/`read_xml()` function in order to manipulate HTML elements. 

```{r}
# xml instance of comparegroups.html
comparegroups.html.parse = read_html(as.character(comparegroups.html))
# xml instance of html_block
plot_row = read_xml(doRenderTags(html_block, indent = F))
```

Then, we look for the first column after `<tbody>` tag because we want to insert our `plot_row` before this first column.

```{r}
# find reference to the first column after tbody
first_tbody_tr = xml_find_first(comparegroups.html.parse, "//tbody//tr")
# add sibling before first_tbody_tr
xml_add_sibling(first_tbody_tr, .value = plot_row, .where = "before")
```

We update our inital table object by replacing the old HTML table with the newest.

```{r}
comparegroups.html[1] = as.character(comparegroups.html.parse)
```


#### Render HTML

To finish, we use `kableExtra` package to generate fashion tables.

```{r, echo=TRUE}
comparegroups.html %>%
  kable_styling(font_size = 12, bootstrap_options = "responsive") %>%
  column_spec(column = 2, bold = T)
```


### Go further

* [Vignette of compareGroups](https://cran.r-project.org/web/packages/compareGroups/vignettes/compareGroups_vignette.html)
* [Vignette of kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
* [DOM elements](https://www.w3schools.com/jsref/dom_obj_all.asp)
* [XPath](https://en.wikipedia.org/wiki/XPath)


